Clustering feature points

### 2013-05-31 text/clustering.tex
Added a bit of text about the test set

\begin{algorithm}
\caption{Mirror Match Algorithm (\emph{MM})}
\label{alg-mm}
\begin{algorithmic}
\Require $images$ : set of images, $threshold$ : float $\in \mathbb{R}$
\State $matches_{init}\gets \varnothing$
\State $matches_{final}\gets \varnothing$
\State $features\gets \varnothing$
\ForAll{$I_i \in images$} \Comment Gather features of all images
	\State $features\gets features \cup getFeatures(I_i)$
\EndFor
\ForAll{$f_i \in features$} \Comment Get best match for all features
	\State $f_m,f_n \gets getTwoNearestNeighbors(f_i, features ~ 
\backslash ~ \left\{f_i\right\})$
	\State $ratio \gets getSimilarity(f_i, f_n) / getSimilarity(f_i, 
f_m)$
	\If{$ratio < threshold$}
		\State $matches_{init}\gets \left(f_i, f_m\right)$
	\EndIf
\EndFor
\ForAll{$\left(f_i, f_j \right) \in matches_{init}$} \Comment Filter 
matches
\If{$\left(f_j, f_i \right) \in matches_{init} \wedge getImg(f_i) \neq 
getImg(f_j)$}
		\State $matches_{final} \gets (f_i, f_j)$
		\State $matches_{init} \gets matches_{init} \ (f_j, f_i)$
	\EndIf
\EndFor \\
\Return $matches_{final}$
\end{algorithmic}
\end{algorithm}

@@

Unlike the \emph{MM} algorithm the feature points are partitioned before 
the matching stage. While the louvain clustering algorithm doesn't 
require any parameters in itself, it tends towards clustering all 
feature points together in the same partition if the graph is fully 
connected or close. To ensure that the graph is well clustered, the 
adjacency matrix is pruned so only edges above a certain threshold is 
kept. From empirical analysis keeping around 2.5\% of edges seems to 
work well. Figure \ref{fig:graph} shows the result of clustering the 
feature points displayed as a graph where each node correspond to a 
feature point. Here the border of nodes signifies the original image 
while the color of the node is decided by what partition it belongs to.  
Notice that since a limited amount of colors are used disconnected 
subgraphs of the same color can be assumed to belong to separate 
partitions. The sub figure \ref{fig:cropped_graph} shows a detailed 
section of the graph where the interplay between small clusters 
consisting of two or three nodes as well as larger clusters consisting 
of several hundred nodes can be seen.

@@

The clusters 
\begin{figure}
	\makebox[\textwidth][c]{%
		\begin{subfigure}[t]{0.60\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/MMC_pitts_source}
			\caption{Source image}
			\label{fig:pitts_source}
		\end{subfigure}%
		%~ %add desired spacing between images, e. g. ~, \quad, \qquad		  
		%(or a blank line to force the subfigure onto a new line)
		\begin{subfigure}[t]{0.60\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/MMC_pitts_keypoints}
			\caption{Keypoints Displayed}
			\label{fig:pitts_keypoints}
		\end{subfigure}%
	}%
	\\
	\makebox[\textwidth][c]{
		%~ %add desired spacing between images, e. g. ~, \quad, \qquad		  
		%(or a blank line to force the subfigure onto a new line)

		\begin{subfigure}[t]{1.16\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/MMC_partition}
			\caption{A Partition Example}
			\label{fig:pitts_partition}
		\end{subfigure}%
	}%
	\label{fig:compare_mirror}
	\caption{Example of partitioning of keypoints in an image}
\end{figure}

\begin{algorithm}
\caption{Impl. of GetMatches (\emph{from MMC algorithm})}
\label{alg-getmatches}
\begin{algorithmic}
\Require $p$ : set of features, $threshold \in \mathbb{R}$, $features$ : 
Set of all features
\State $matches \gets \varnothing$
\State $edges \gets \left\{getSimilarity(f_i, f_j) \mid getImg(f_i) \neq 
getImg(f_j) \wedge f_i, f_j \in p \right\}$
\If{$\left\vert edges \right\vert = 1$} \Comment If $\exists$ one edge 
between images
	\State $f_i,f_j \gets getInterImageFeatures(edges_{inter}, p)$
	\State $f_m,f_n \gets getTwoNearestNeighbors(f_i, features ~ 
\backslash ~ \left\{f_i\right\})$
	\State $f_s,f_t \gets getTwoNearestNeighbors(f_j, features ~ 
\backslash ~ \left\{f_j\right\})$
	\State $ratio_i \gets getSimilarity(f_i, f_n) / getSimilarity(f_i, 
f_j)$
	\State $ratio_j \gets getSimilarity(f_j, f_t) / getSimilarity(f_j, 
f_i)$
	\If{$ratio_i < threshold \wedge ratio_j < threshold$}
		\State $matches \gets matches \cup (f_i, f_j)$
	\EndIf
\ElsIf{$\left\vert edges \right\vert > 1$} \Comment If $\exists$ more 
than one edge between images
	\ForAll{$f_i \in p$}
		\State $f_m,f_n \gets getTwoNearestNeighbors(f_i, p ~ \backslash 
~ \left\{f_i\right\})$
		\State $ratio \gets getSimilarity(f_i, f_n) / getSimilarity(f_i, 
f_m)$
		\If{$getImg(f_i) \neq getImg(f_m) \wedge ratio < threshold 
\wedge (f_m, f_i) \not\in matches$}
			\State $matches \gets matches \cup (f_i, f_m)$
		\EndIf
	\EndFor
\EndIf

@@

Algorithm \ref{alg-mmc} shows the basic steps taken to find matches in 
two or more images.

This algorithm is a simplified version of the proposed method completely 
agnostic of the geometry of the feature points in the picture. To better 
explain the algorithm here is a step by step explication of the inputs 
and functions mentioned:
\begin{itemize}
\item[]{\emph{images}: The images we are matching}
\item[]{\emph{$n$}: The number of weights per row in the adjecency 
matrix.}
\item[]{\emph{getFeatures}: Returns a set of feature points given an 
image. In practice the BRIEF\cite{calonder2010brief} feature point 
detector and descriptor has worked well, but it could easily be replaced 
with any other feature point type.} 
% TODO: include a comparison of SURF, SIFT, etc
\item[]{\emph{getAdjecencyMatrix}: Based on the distance measure used to 
compare the similarity of the feature points a symmetric adjecency 
matrix is returned.}
\item[]{\emph{pruneEdges}: The result of the Louvain clustering depends 
on the density of the graph. To ensure an optimal number of partitions 
we only keep the $n$ highest weights per row and column in the adjecency 
matrix, making sure it's still symmetric afterwards. In practice setting 
$n$ to around four or five yields good results.}
\item[]{\emph{cluster}: Cluster the graph corresponding to the adjecency 
matrix given using the Louvain algorithm in \cite{blondel2008}.}
\item[]{\emph{getPositions}: Given a trait consisting of features in 
several different images, find the positions of traits in each image and 
return them. This could either be done by taking a mean of the feature 
points in an image, or by returning the position of the highest matching 
feature points within the trait. In practice ... TODO}
\end{itemize}

@@

\section{Experiments and Image Set}
\label{experiment}

\section{Results}
\label{results}
\subsection{Graph Clustering}

\subsubsection{On the subject of features}

TODO: Show examples of the same image with feature points from different 
detectors and explain the difference. Go in to why brief seems to work 
best in practice and superficially explain how brief works compared with 
SIFT.

\begin{algorithm}
\caption{Cluster-Match Algorithm With Geometry}
\label{alg-geometry}
\begin{algorithmic}
\Require $images$ : set of images, $\alpha$ : float $\in \left[0, 
1\right]$, $\beta$ : float $\in \left[0, 1\right]$
\State $matches\gets \varnothing$
\ForAll{$I_i \in images$} \Comment Gather features of all images
	\State $f_i\gets getFeatures(I_i)$
\EndFor
\State $A\gets getAdjecencyMatrix(f_1, f_2,\; \ldots \;, f_n)$
\State $B\gets getGeomDistanceMatrix(f_1, f_2,\; \ldots \;, f_n)$
\State $M\gets getInterImageMask(f_1, f_2,\; \ldots \;, f_n)$
\State $B_{norm}\gets 1 - normalize(G)$ \State $A_{norm}\gets 1 - 
normalize(A)$
\State $A_{geom}\gets A_{norm}\left[ \neg M\right] + \beta \cdot 
B_{norm}\left[M\right]$
\State $A_{pruned}\gets pruneEdges(A_{geom},\alpha)$
\State $P\gets cluster(A_{pruned})$ 
\ForAll{$p \in P$}
	\State $trait\gets \left\{p \cap f_i \mid p \cap f_i \neq 
\varnothing, f_i \in \left\{f_1, f_2,\; \ldots \;, f_n\right\}\right\}$
	\If{$\left\vert trait \right\vert > 1$}
		\State $matches\gets getPositions(trait)$
		\Comment Position of trait in each image
	\EndIf
\EndFor \\
\Return matches
\end{algorithmic}
\end{algorithm}

While clustered feature points are often geometrically close together, 
there might be cases where several points in the same image are similar 
enough to all be included in one cluster despite their geometrical 
distance.  To solve this problem we can either modify the clustering 
algorithm to take the geometric distances into account like 
\cite{das2008event}.  Alternatively we can modify the weight matrix to 
penalize putting distanced feature points in the same cluster. However 
we only know the geometrical distance between two feature points as long 
as both belong to the same image. One way to approach this problem is to 
replace all the weights of edges between two feature points within the 
same image with their distance. For the case of two images the 
normalized adjecency matrix is defined as follows:
%
$$A_{norm}=
\begin{bmatrix}
w_{1,1}   & w_{1,2}   & \ldots & w_{1,n-1}   & w_{1,n}     \\
w_{2,1}   & w_{2,2}   & \ldots & w_{2,n-1}   & w_{2,n}     \\
\vdots    & \vdots    & \ddots & \vdots      & \vdots      \\
w_{n-1,1} & w_{n-1,2} & \ldots & w_{n-1,n-1} & w_{n-1,n}   \\
w_{n,1}   & w_{n,2}   & \ldots & w_{n,n-1}   & w_{n,n}
\end{bmatrix}=
\begin{bmatrix}
A_{im1} & S       \\
T       & A_{im2} \\
\end{bmatrix}
$$
%
In a similar manner we can define an adjecency matrix based on the 
geometrical distance. Note for example how $d_{1,n}$ can take any value 
as it denotes the geometric distance of feature points in two different 
images and as such can be set to any value.
%
$$B_{norm}=
\begin{bmatrix}
d_{1,1}   & d_{1,2}   & \ldots & d_{1,n-1}   & d_{1,n}     \\
d_{2,1}   & d_{2,2}   & \ldots & d_{2,n-1}   & d_{2,n}     \\
\vdots    & \vdots    & \ddots & \vdots      & \vdots      \\
d_{n-1,1} & d_{n-1,2} & \ldots & d_{n-1,n-1} & d_{n-1,n}   \\
d_{n,1}   & d_{n,2}   & \ldots & d_{n,n-1}   & d_{n,n}
\end{bmatrix}=
\begin{bmatrix}
B_{im1} & R       \\
Q       & B_{im2} \\
\end{bmatrix}
$$
%
Given these two and a weight factor $\beta$ we can construct a new 
weight matrix as follows:
%
$$C=
\begin{bmatrix}
\beta \cdot B_{im1} & S       \\
T       & \beta \cdot B_{im2} \\
\end{bmatrix}
$$
%
Since we prune the matrix afterwards, it's important to choose a weight 
factor that doesn't lead to all edges being either intra image edges or 
intra image edges, since the first would create no clusters across 
images and the latter would mean that the geometrical information would 
be discarded and the bipartite graph constituting of edges between 
feature points in the two images would be used instead.

>>>>

\begin{algorithm}
\caption{Mirror Match Algorithm (\emph{MM})}
\label{alg-mm}
\begin{algorithmic}
\Require $images$ : set of images, $threshold$ : float $\in \mathbb{R}$
\State $matches_{init}\gets \varnothing$
\State $matches_{final}\gets \varnothing$
\State $features\gets \varnothing$
\ForAll{$I_i \in images$} \Comment Gather features of all images
	\State $features\gets features \cup getFeatures(I_i)$
\EndFor
\ForAll{$f_i \in features$} \Comment Get best match for all features
	\State $f_m,f_n \gets getTwoNearestNeighbors(f_i, features ~ 
\backslash ~ \left\{f_i\right\})$
	\State $ratio \gets getSimilarity(f_i, f_n) / getSimilarity(f_i, 
f_m)$
	\If{$ratio < threshold$}
		\State $matches_{init} \gets \left(f_i, f_m\right)$
	\EndIf
\EndFor
\ForAll{$\left(f_i, f_j \right) \in matches_{init}$} \Comment Filter 
matches
\If{$\left(f_j, f_i \right) \in matches_{init} \wedge getImg(f_i) \neq 
getImg(f_j) \wedge \left(f_j, f_i\right) \not\in matches_{final}$}
		\State $matches_{final} \gets (f_i, f_j)$
	\EndIf
\EndFor \\
\Return $matches_{final}$
\end{algorithmic}
\end{algorithm}

@@

Unlike the \emph{MM} algorithm the feature points are partitioned before 
the matching stage. While the louvain clustering algorithm doesn't 
require any parameters in itself, it tends towards clustering all 
feature points together in the same partition if the graph is fully 
connected or close. To ensure that the graph is well clustered, the 
adjacency matrix is pruned so only edges above a certain threshold is 
kept. From empirical analysis keeping around 2.5\% of edges seems to 
work well. Figure \ref{fig:graph} shows the result of clustering the 
feature points displayed as a graph where each node correspond to a 
feature point. Here the border of nodes signifies the original image 
while the color of the node is decided by what partition it belongs to.  
Notice that since a limited amount of colors are used disconnected 
subgraphs of the same color can be assumed to belong to separate 
partitions. The sub figure \ref{fig:cropped_graph} shows a detailed 
section of the graph where the interplay between small partitions 
consisting of two or three nodes as well as larger partitions consisting 
of several hundred nodes can be seen.

@@

The partitions group feature points by similarity which means that 
repetitive structures such as buildings often feature in larger 
partitions as exemplified in figure \ref{fig:compare_mirror}. Here the 
cluster shown in part \ref{fig:pitts_partition} contains left window 
corners from two buildings across both images. This is an example of 
feature points that are hard to match correctly since each point will 
have several similar potential matches.

\begin{figure}
	\makebox[\textwidth][c]{%
		\begin{subfigure}[t]{0.60\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/MMC_pitts_source}
			\caption{Source image}
			\label{fig:pitts_source}
		\end{subfigure}%
		%~ %add desired spacing between images, e. g. ~, \quad, \qquad		  
		%(or a blank line to force the subfigure onto a new line)
		\begin{subfigure}[t]{0.60\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/MMC_pitts_keypoints}
			\caption{Keypoints Displayed}
			\label{fig:pitts_keypoints}
		\end{subfigure}%
	}%
	\\
	\makebox[\textwidth][c]{
		%~ %add desired spacing between images, e. g. ~, \quad, \qquad		  
		%(or a blank line to force the subfigure onto a new line)

		\begin{subfigure}[t]{1.16\textwidth}
			\centering
			\includegraphics[width=\textwidth]{images/MMC_partition}
			\caption{A Partition Example}
			\label{fig:pitts_partition}
		\end{subfigure}%
	}%
	\label{fig:compare_mirror}
	\caption{Example of a partition of keypoints in an image}
\end{figure}

The matching algorithm for \emph{MMC} (algorithm \ref{alg-getmatches}) 
goes through the partitions one by one and selects matches in the 
following way: If the partition has no edges going from a feature point 
in one image to a feature point in another image, the partition is 
discarded. If we have more than one edge going between images then 
$getMatches$ will behave as \emph{MM} and make sure that a 
correspondence is symmetric, that it has a ratio above the desired 
threshold and that it isn't matching two feature points from the same 
image. Since feature points are all guaranteed to be fairly similar this 
means that we can set the ratio threshold much higher than in \emph{MM} 
and include matches that would normally have been discarded for not 
being sufficiently unique.
However as can be seen in the example in figure \ref{fig:cropped_graph} 
many of the partitions contain only two feature points from different 
images linked by one edge. This is often the case if all edges below a 
certain similarity threshold have been removed in the graph, in which 
case it's probable that some nodes will end up with only one outgoing 
edge. This doesn't necessarily imply that the correspondence represented 
by the edge is unique, but the only way to figure out if it is, is by 
checking the similarity scores of the nearest neighbor as proposed in 
\cite{lowe2004sift}.

\begin{algorithm}
\caption{Impl. of getMatches (\emph{from MMC algorithm})}
\label{alg-getmatches}
\begin{algorithmic}
\Require $p$ : set of features, $threshold \in \mathbb{R}$, $features$ : 
Set of all features
\State $matches \gets \varnothing$
\State $edges \gets \left\{getSimilarity(f_i, f_j) \mid getImg(f_i) \neq 
getImg(f_j) \wedge f_i, f_j \in p \right\}$
\If{$\left\vert edges \right\vert = 1$} \Comment If $\exists$ one edge 
between images
	\State $f_i,f_j \gets getInterImageFeatures(edges_{inter}, p)$
	\State $f_m,f_n \gets getTwoNearestNeighbors(f_i, features ~ 
\backslash ~ \left\{f_i\right\})$
	\State $f_s,f_t \gets getTwoNearestNeighbors(f_j, features ~ 
\backslash ~ \left\{f_j\right\})$
	\State $ratio_i \gets getSimilarity(f_i, f_n) / getSimilarity(f_i, 
f_j)$
	\State $ratio_j \gets getSimilarity(f_j, f_t) / getSimilarity(f_j, 
f_i)$
	\If{$ratio_i < threshold \wedge ratio_j < threshold$}
		\State $matches \gets matches \cup (f_i, f_j)$
	\EndIf
\ElsIf{$\left\vert edges \right\vert > 1$} \Comment If $\exists$ more 
than one edge between images
	\ForAll{$f_i \in p$}
		\State $f_m,f_n \gets getTwoNearestNeighbors(f_i, p ~ \backslash 
~ \left\{f_i\right\})$
		\State $ratio \gets getSimilarity(f_i, f_n) / getSimilarity(f_i, 
f_m)$
		\If{$getImg(f_i) \neq getImg(f_m) \wedge ratio < threshold 
\wedge (f_m, f_i) \not\in matches$}
			\State $matches \gets matches \cup (f_i, f_m)$
		\EndIf
	\EndFor
\EndIf

@@

@@

\section{Experimental Setup}
\label{experiment}

To reliably measure the accuracity of a matching method on real images 
we either need a set of image pairs tied by a homography or manually 
count the number of inliers. The latter poses becomes unfeasable the 
moment we attempt to test a non-trivial set of images, but has the 
advantage of working with any image pair. Using a homography between 
image pairs reduces the test cases to images where the matched areas lie 
on a plane, or alternatively that there is no viewpoint change. This 
greatly reduces the body of image pairs we can use to reliably test 
matching methods. In \cite{mikolaczyk2005performance} Mikolajczyk and 
Schmid introduces a set of images to compare the performance of feature 
detectors. The test set looks at different feature descriptors 
invariance faced with different lighting, blur, rotation as well as 
viewpoint change. However in practice the image used for the view point 
change featuring a wall of graffiti\footnote{To ensure that all points 
lie on the same plane} has gone on to be used the most for later tests 
given that most of the other test pairs have since become relatively 
trivial\footnote{see \cite{wu2011robust} or \cite{delponte2006svd} for 
examples}.

Inspired by this I have collected a set of image pairs featuring murals 
of a few famous artists\footnote{Including Banksy, Blu and Shepard 
Fairey among others}. Each pair consist of images of the same motif 
taken from different angles, often including repetitive patterns and 
texture. Based on this image set the two algorithms are tested on a 
series of image patches from each pair. In order to test if a matching 
is reliable we of course need correct matches when the image pair we are 
testing on indeed features the same scene, but we also need to make sure 
that we don't get any matches when the two images have nothing in 
common. This means that it isn't enough just testing against pairs of 
images that we know match. We also need to test against pairs that might 
look like they could match but in fact don't.

With this in mind I have created a test suite by cropping out $300$ 
pixel by $300$ pixel patches of the original pair with a random vertical 
and horizontal offset. Given an image pair I can obtain $n$ pairs of 
patches where the two patches in any pair might or might not overlap.  
This ensures that the patch pairs where there is no overlap still retain 
a basic similarity to each other, while the patches that do overlap 
often only overlap on a small part of their surface.

\section{Results}
\label{results}
